Functors, CPS and monads, or how to generate efficient code from abstract designs

### 2005-04-06 paper/metamonads.tex
making text more concise

\usepackage{hyphenat}
\usepackage{comment}
\usepackage{amsmath,amssymb}
\usepackage{amstext}
\usepackage{url}

@@

\usepackage{fancyvrb}
\DefineShortVerb{\|}
\DefineVerbatimEnvironment{code}{Verbatim}{xleftmargin=\mathindent}
\setlength{\parskip}{0pt}
\newlength{\mathindent}
\setlength{\mathindent}{1em}

@@

The rest of this paper is structured as follows:
The next section introduces code generation in MetaOCaml, the problem
of name generation, and how continuation-passing style is a general
tool for solving that problem.  Section~\ref{monadicnotation} introduces
the monad that greatly simplifies the task of dealing with
possibly effect-inducing CPS code combinators.  Section~\ref{functors}
shows how one can use the advanced module system of OCaml, namely its
parametrized modules, to encode all of the aspects of the Gaussian
Elimination algorithm family in completely separate, independent modules.
We then briefly discuss related work in section~\ref{related}, outline
some work still to do in section~\ref{future}, and finally draw some 
conclusions in section~\ref{conclusion}.  As an appendix, we show
the details of the syntax extension for monadic programming in Ocaml
and MetaOCaml.

@@

@@

One particular instantiation of the domain (a particular structure of
|DOMAIN| signature) is the |int| domain \oleg{should we put the following?}
|IntegerDomain : DOMAIN| to say that the compiler will verify that our
|IntegerDomain| is indeed a |DOMAIN|, that is, satisfies the required
signature. The constraint |DOMAIN| may be omitted; in that case, 
the compiler will verify the type when we try to use that structure as
a |DOMAIN|. In any case, the errors such as missing ``methods'' or 
methods of wrong types will be caught statically, even \emph{before} any code
generation takes place.
\begin{small}
\begin{code}
module IntegerDomain = struct
    type v = int
    type kind = domain_is_ring
    type 'a vc = ('a,v) code
    let zero = .< 0 >.  
    let one = .< 1 >. 
    let plus x y = ret .<.~x + .~y>. 
    let times x y = ret .<.~x * .~y>.
    let minus x y = ret .<.~x - .~y>.
    let uminus x = ret .< - .~x>.
    let div x y = ret .<.~x / .~y>. 
    let better_than = Some (fun x y -> ret .<abs .~x > abs .~y >. )
    let normalizerf = None 
    let normalizerg = fun x -> x
end
\end{code}
\end{small}
\noindent While the |DOMAIN| type may have looked daunting to some, 
the implementation is quite straightforward.  It is equally simple to 
make implementations for |float| and for |Num.num| (arbitrary precision
exact rational numbers).  With scarcely more work (as these are not 
built-in and the arithmetic must be implemented), $\mathbb{Z}_p$,
$\mathbb{Z}\left[x\right]$, $\mathbb{Q}\left(x\right)$ and more
complex domains can be done as well.

\begin{small}
\begin{code}
module type DETERMINANT = sig
  type indet
  type outdet
  type tdet = outdet ref
  type 'a lstate
  val decl : unit -> 
    (unit, [> `TDet of 'a lstate ] list, ('a,'b) code) monad
  val upd_sign : unit -> 
    (('a,unit) code, [> `TDet of 'a lstate ] list, ('a,'b) code) monad
  val zero_sign : unit -> 
    (('a,unit) code, [> `TDet of 'a lstate ] list, ('a,'b) code) monad
  val acc : ('a,indet) code -> 
    (('a,unit) code, [> `TDet of 'a lstate ] list, ('a,'b) code) monad
  val get : unit ->
    (('a,tdet) code, [> `TDet of 'a lstate ] list, ('a,'b) code) monad
  val set : ('a,indet) code -> 
    (('a,unit) code, [> `TDet of 'a lstate ] list, ('a,'b) code) monad
  val fin : unit -> 
    (('a,outdet) code, [> `TDet of 'a lstate ] list, ('a,'b) code) monad
end
\end{code}
\end{small}
\noindent  Determinant tracking is quite complex: we need to 
declare that we are tracking the determinant (|decl|), have some functions
for tracking the sign of the determinant (|upd_sign|,|zero_sign|),
|set|, |get|, accumulate (|acc|) unsigned values for the determinant, as 
well as returning a final value (|fin|) for the determinant.  In some cases,
like for float matrices where we do not return the determinant, all of
the above should generate no code at all, while for integer matrices where
we want the determinant as part of the output, each of these need to be
implemented.

\bibliography{metamonads}
\bibliographystyle{plain}

Perhaps give the command line how to load the camlp4 module into
ocaml. Point to the code.
\end{document}

>>>>

\usepackage{hyphenat}
\usepackage{comment}
\usepackage{amsmath,amssymb}
\usepackage{amstext}
\usepackage{url}
\usepackage[dvips]{color}

@@

\usepackage{fancyvrb}
\DefineShortVerb{\|}
\DefineVerbatimEnvironment{code}{Verbatim}{xleftmargin=\mathindent,commandchars=\\\{\}}
\newcommand{\evalresult}[1]{\ensuremath{\Longrightarrow}\textcolor{red}{#1}}
\setlength{\parskip}{0pt}
\newlength{\mathindent}
\setlength{\mathindent}{1em}

@@

The rest of this paper is structured as follows:
The next section introduces code generation in MetaOCaml, the problem
of name generation, and how continuation-passing style is a general
tool for solving that problem.  We also introduce
the monad that greatly simplifies the task of dealing with
possibly effect-inducing CPS code combinators.  Section~\ref{functors}
shows how one can use the advanced module system of OCaml, namely its
parametrized modules, to encode all of the aspects of the Gaussian
Elimination algorithm family in completely separate, independent modules.
We then briefly discuss related work in section~\ref{related}, outline
some work still to do in section~\ref{future}, and finally draw some 
conclusions in section~\ref{conclusion}.  As an appendix, we show
the details of the syntax extension for monadic programming in Ocaml
and MetaOCaml.

@@

\section{The problem of generating names, CPS, and monad}\label{CPS}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% And now, a better -- and shorter -- write-up.

We will be using MetaOcaml which, as an instance of a multi-stage
programming system \cite{TahaThesis}, provides exactly the needed
features: to construct a code expression, to combine them, and to
execute them. The figure XXX shows the simplest code generator |one|,
and more complex generators.

\begin{figure*}
\begin{code}
let one = .<1>. and plus x y = .<.~x + .~y>.
let simplest_code = let gen x y = plus x (plus y one) in
  .<fun x y -> .~(gen .<x>. .<y>.)>.
\evalresult{.<fun x_1 -> fun y_2 -> (x_1 + (y_2 + 1))>.}
let simplest_param_code plus one = let gen x y = plus x (plus y one) in
  .<fun x y -> .~(gen .<x>. .<y>.)>.
let plus x y = .<.~x +. .~y>. and one = .<1.0>. in
  simplest_param_code plus one
let param_code1 plus one =
  let gen x y = plus (plus y one) (plus x (plus y one)) in
  .<fun x y -> .~(gen .<x>. .<y>.)>.
let param_code1' plus one =
  let gen x y = let ce = (plus y one) in  plus ce (plus x ce) in
  .<fun x y -> .~(gen .<x>. .<y>.)>.
param_code1' plus one
\evalresult{.<fun x_1 -> fun y_2 -> ((y_2 + 1) + (x_1 + (y_2 + 1)))>.}
\end{code}
\caption{Code generation and combinators. $\Longrightarrow$ shows the
  result of the evaluation of the preceeding expression}
\end{figure*}

We use MetaOCaml brackets |.<...>.| to generate code expressions -- in
other words, to construct future-stage computations. We use escapes
|.~| to perform an immediate code generating computation while we are
building the future-stage computation (code expression). The immediate
computation in |simplest_code| includes the evaluation of the function
|gen|, which in turn applies the function |plus|. The function
receives code expressions |.<x>.| and |.<y>.| as arguments. At the
generating stage, we can manipulate code expressions as (opaque)
values. The function |gen| returns a code expression, which will be
inlined in the place of the escape. MetaOCaml can print out code
expressions, so we can see the final generated code, It has
no traces of the functions |gen| and |plus|: their applications are
done at the code generation stage.

The final MetaOCaml feature, |.!| (pronounced ``run'') 
executes the code expression: |.! simplest_code| is a function of two
integers, which we can apply: |(.! simplest_code) 1 2|. The original
|simplest_code| is not the function on integers -- it is a code
expression.

To see the benefit of code generation, we notice that we can easily
parameterize our code, |simplest_param_code|, and use it to generate
code that operates on integers, floating point numbers or booleans --
in general, any domain that implements |plus| and |one|.

A more complex expression |param_code1| has two occurrences of 
|plus y one|. Depending on the implementation of |plus| that may be quite a
complex computation, and so we would rather not do it twice. We may be
tempted to rely on the compiler's common-subexpression elimination
optimization. We would rather not rely on a ``sufficiently smart
compiler'': when the generated code is very complex, the compiler may
overlook common subexpressions.  Or the subexpressions may occur in an
imperative context where the compiler might not be able to determine
if lifting them is sound. So, being conservative, the optimizer will
leave the duplicates as they are. We may attempt to eliminate
subexpressions as in |param_code1'|. However, the result of
|param_code1' plus one| still exhibits duplicate sub-expressions. 
Our |let|-insertion
optimization saved the computation at the generating stage. 
We need a combinator that inserts the |let| expression in the
generat\emph{ed} code. We need a combinator |letgen| to be used
as\\|let ce = letgen (plus y one) in plus ce (plus x ce)| 
yielding the code like |.<let t = y + 1 in t + (x + t)>.|
But that seems impossible because |letgen exp| has to generate
the expression\\|.<let t = exp in body>.| although |letgen| does not have
the |body| yet. The body needs a temporary identifier |.<t>.| that
is supposed to be the result of |letgen| itself. 
Certainly |letgen| cannot generate only part of a let-expression,
without the |body|,  as all generated expressions in
MetaOCaml are well-formed and complete.

The key is to use continuation-passing style (CPS). Its benefits were first
pointed out by \cite{Bondorf:92} in the context
of partial evaluation, and extensively used by \cite{KiselyovTaha} for code
generation. Now, |param_code2 plus one| gives us the desired code.

\begin{small}
\begin{code}
let letgen exp k = .<let t = .~exp in .~(k .<t>.)>.
let param_code2 plus one =
  let gen x y k = letgen (plus y one) (fun ce -> k (plus ce (plus x ce)))
  and k0 x = x
  in .<fun x y -> .~(gen .<x>. .<y>. k0)>.
param_code2 plus one
\evalresult{.<fun x_1 -> fun y_2 -> let t_3 = (y_2 + 1) in (t_3 + (x_1 + t_3))>.}
\end{code}
\end{small}

Finally, |runM| runs our monad, that is, performs the computation of
the monad and gets the result, which in our case is the code
expression. We run the monad by passing it the initial state and the
initial continuation |k0|. We can now re-write our |param_code2|
example of the previous section as |param_code3|, see Fig. XXX.

\begin{figure*}
\begin{code}
let param_code3 plus one =
  let gen x y = bind (retN (plus y one)) (fun ce -> 
                ret (plus ce (plus x ce)))
  in .<fun x y -> .~(runM (gen .<x>. .<y>.))>.
let param_code4 plus one =
  let gen x y = mdo \{ ce <-- retN (plus y one);
                      ret (plus ce (plus x ce)) \}
  in .<fun x y -> .~(runM (gen .<x>. .<y>.))>.
let ifM' test th el = mdo \{
  testc <-- test; thc <-- th; elc <-- el;
  ifL testc thc elc\}
let gen a i = ifM' (ret .<(.~i) >= 0>.) (retN .<Some (.~a).(.~i)>.) (ret .<None>.)
 in .<fun a i -> .~(runM (gen .<a>. .<i>.))>.
\evalresult{.<fun a_1 -> fun i_2 -> 
    let t_3 = (Some (a_1.(i_2))) in if (i_2 >= 0) then t_3 else (None)>.}
let gen a i = ifM (ret .<(.~i) >= 0>.) (retN .<Some (.~a).(.~i)>.) (ret .<None>.)
 in .<fun a i -> .~(runM (gen .<a>. .<i>.))>.
\evalresult{.<fun a_1 -> fun i_2 ->
    if (i_2 >= 0) then let t_3 = (Some (a_1.(i_2))) in t_3 else (None)>.}
\end{code}
\end{figure*}
% param_code3 plus one;;

That does not seem like much of an improvement. With the help of
camplp4 pre-processor (cite), we introduce the do notation (give URL
or cite the message on the MetaOCaml list), patterned after the
do-notation of Haskell (cite). The function |param_code4|, written in
the do-notation, is equivalent to |param_code3| -- in fact, the camlp4
preprocessor will convert the former into the latter. And yet,
|param_code4| looks far more conventional, as if it were indeed in
direct style.

We can write operations that generate code other than let-statements,
e.g., conditionals: see |ifL| in Figure~\ref{ourmonad}. The function |ifL|, 
albeit straightforward, is not as general as we wish: its argument are
already generated pieces of code rather than monadic values. So, we
may want to ``lift it'', see |ifM'|. We can define functions |l1|,
|l2|, |l3| (analogues of |liftM|, |liftM2|, |liftM3| of Haskell) 
to make such a lifting generic.
We soon notice the need for another |ifM| function, with the same
interface (see |ifM| in Figure~\ref{ourmonad}). The difference between them is
apparent, see Fig. XXX: in the code with |ifM'|, the let-insertion
happend \emph{before} the if-expression, that is, before the test that
the index |i| is positive. That is, if |i| turned out
negative, |a.(i)| will generate an out-of-bound array access
error. On the other hand, the code with |ifM| accesses the array only
when we have verified that the index is non-negative. This example
makes it clear that the code generation (such as the one in |retN|) is 
truly an effect and we have to be clear about the sequencing of
effects, when generating control constructions such as conditionals.
The form |ifM| handles the such effects correctly. In our code, we
need similar operators for other Ocaml control forms: for generating
case-matching statements and for- and while-loops.

One particular instantiation of the domain (a particular structure of
|DOMAIN| signature) is the |int| domain \oleg{should we put the following?}
|IntegerDomain : DOMAIN| to say that the compiler will verify that our
|IntegerDomain| is indeed a |DOMAIN|, that is, satisfies the required
signature. The constraint |DOMAIN| may be omitted; in that case, 
the compiler will verify the type when we try to use that structure as
a |DOMAIN|. In any case, the errors such as missing ``methods'' or 
methods of wrong types will be caught statically, even \emph{before} any code
generation takes place.
\begin{small}
\begin{code}
module IntegerDomain : DOMAIN = struct
  type v = int  type kind = domain_is_ring
  type 'a vc = ('a,v) code
  let zero = .< 0 >.  and one = .< 1 >. 
  let plus x y = ret .<.~x + .~y>. 
  let better_than = Some (fun x y -> ret .<abs .~x > abs .~y >. )
  let normalizerf = None
  ...
end
\end{code}
\end{small}
\noindent Here |domain_is_ring| is an abstract data type to encode semantic constraint
-- the kind of the domain.
While the |DOMAIN| type may have looked daunting to some, 
the implementation is quite straightforward.  It is equally simple to 
make implementations for |float| and for |Num.num| (arbitrary precision
exact rational numbers).  With scarcely more work (as these are not 
built-in and the arithmetic must be implemented), $\mathbb{Z}_p$,
$\mathbb{Z}\left[x\right]$, $\mathbb{Q}\left(x\right)$ and more
complex domains can be done as well.

\begin{small}
\begin{code}
module type DETERMINANT = sig
  type indet  type outdet  type 'a lstate
  type tdet = outdet ref   
  val decl : unit -> 
    (unit, [> `TDet of 'a lstate ] list, ('a,'b) code) monad
  val upd_sign : unit -> 
    (('a,unit) code, [> `TDet of 'a lstate ] list, ('a,'b) code) monad
  ...
end
\end{code}
\end{small}
\noindent  Determinant tracking is quite complex: we need to 
declare that we are tracking the determinant (|decl|), have some functions
for tracking the sign of the determinant (|upd_sign|,|zero_sign|),
|set|, |get|, accumulate (|acc|) unsigned values for the determinant, as 
well as returning a final value (|fin|) for the determinant.  In some cases,
like for float matrices where we do not return the determinant, all of
the above should generate no code at all, while for integer matrices where
we want the determinant as part of the output, each of these need to be
implemented.

\bibliography{metamonads}
\bibliographystyle{plain}
\end{document}

% It seems no appendices are allowed. Check with Walid?

Perhaps give the command line how to load the camlp4 module into
ocaml. Point to the code.

%%% Old, detailed version is here

@@
